Testing and Isolation Efficacy; Insights from a Simple Epidemic Model

### 2020-12-08 SIR_manuscript.tex
working on ms

\jd{There is a gap here between unravelling data and what we actually do.}

@@

It is often the purpose of testing, apart from the testing infrastructure, that dictates/defines \ali{[better word?]} the testing strategy. There are at least four different, and sometimes conflicting, purposes of testing: (i) diagnosis and treatment, (ii) screening people for access to things like flights, surgery, LTCs etc, (iii) surveillance, and (iv) mitigation and control of disease spread (ref to Troy Gayle?). Further, the testing strategies can be grouped into two categories including random testing across the whole population and Non-random/biosed testing. Random testing is the simplest form of testing strategy and is typically carried out for the purpose of surveillance. It is notable that, by random testing we mainly refer to PCR test given that there is almost negligable virological testing for survilance purposes. Furthermore, all of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are non-random testing strategies with a different bios. Specifically, when the purpose is "(i) diagnosis and treatment" or "(iv) mitigation and control of disease spread", the bios is towards infectious individuals more strongly for (i) than (iv). When people are tested for "(ii) screening", the bios is towards the group-specific charactristics [I am making this up for now!]. For example, people who are getting on flights may/will be more mobile than people awaiting for a surgery and are probably going to stay in long-term care homes and consequently more isolated to begin with.

-Examples of random versus biosed testing strategies;
Each of these strategies has been used as a tool to study COVID-19; see (refs) for group (i) and (refs) for group (ii). The advantage of non-random testing comparing to random testing has been investigated specifically during the COVID-19 pandemic (refs).

@@

We developed a deterministic model, \crefrange{eq1}{eq12}, which groups individuals based on disease status and testing status. The former includes Susceptible, Infectious and Recovered, thus SIR, and the latter categorizes people as \emph untested, waiting-for-\emph positive, waiting-for-\emph negative, or \emph confirmed positive. Also, two 'accumulator' compartments, $N$ and $P$, were incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of $\Rnum$ are presented in the Supplimentary Materials. The flowchart model is as follows

@@

The classical SIR model is based on the following implicit assumptions (see a ref for details); well-mixed population, homogeneity of the population, exponentially distributed duration of infection and large population size (may be refer the reader to \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \crefrange{eq1}{eq12}, assumes: (i) there is a single force of infection, $\Lambda$, that is, where a susceptible individual moves to does not depend on who infected it, (ii) $\eta_c \leq \eta_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals, (iii) a perfectly specific test, $p_S=0$, for simplicity in analysis. Note that the latter assumption combined with the assumption that the total population was initially distributed between untested and awaiting for negative test, that is $S_u+S_n=0$, reduce the model to 10 equations with equations \ref{eq3} and \ref{eq4} are eliminated. Specifically, when the population is at steady states and the number of infected people is negligable, the assumption of perfect specificity results in $S_p=0$ in \cref{eq3} and $d S_c/dt=0$ from \cref{eq4}. Also, since we assumed $S_u+S_n=0$, initially $S_c=0$ and stays in this state.

\newgeometry{margin=1cm} % modify this if you need even more space
\begin{landscape}
\begin{table}[htp]
\centering
{\tiny %
\begin{tabular}{|c|ccc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of onward flow from the awaiting to reported or untested compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & Per capita testing intensity & 1/day & 0.01 \\ \hline 
  $\eta_w$  & Relative probability of transmission for isolated awaiting individuals & - & - \\ \hline
  $\eta_c$  & Relative probability of transmission for isolated confirmed individuals & - & -  \\ \hline
  $\Lambda$ & Force of infection & 1/time & - \\ \hline
  $p_S$ & Probability of false positive for susciptibles& - & 0 \\ \hline
  $p_I$ & Probability of being infected and tested positive & - & 1 \\ \hline
  $p_R$ & Probability of being recovered and tested positive & - & 0.5 \\ \hline
  $W_S, W_I, W_R$ & Relative testing weight & - & 
  \begin{minipage}[t]{0.21\columnwidth}%
 Random testing: $W_S=W_I=W_R=1$, \\Non-random testing: $W_S=0.3, W_I=W_R=1$
\end{minipage} \\ \hline

@@

\todo{Something needs to be said here about simulation part.} The analytical calculation of the next generation matrix, $G$, and the derivation of $\Rnum$ was carried out in Maple (ref) by using simple linear Algebra package. 
  We used \Rlogo for the simulation part which included using the explicit expression of $\Rnum$ \ref{R0} as a function of the underlying parameters, specifying the parameters a realistic range, calculating the corresponding $\Rnum$ and plotting the contours of $\Rnum$ by using ggplot package in \Rlogo. The related result is presented in Figure \ref{pan}, where  panel \eqref{p.a} represents the random testing, and panel \eqref{p.b} is representing non-random testing. The parameter values or ranges are as follows (I like using a table here, with the param, their units and values). It is notable that in the simulation part, the explicit expression of $\Rnum$ was used and not the approximation \eqref{eq:R0appr}. This simulation reflects the behavior of $\Rnum$ with respect to the selected parameters for two different testing strategies: (i) random testing, represented by all testing weights to be the same, $W_S=W_I=W_R$, and (ii) non-random testing, when testing weight are not equal. For simmulation purposes we chose $W_S=W_I=W_R=1$ for random testing, and $W_S=0.3$ and $W_I=W_R=1$ for non-random testing strategy. Note that the critical contour of $\Rnum=1$ is plotted in solid line in Figure \ref{pan}.

@@

Our numerical simulation in support of the analytical results in Proposition \ref{prop1} are presented in Figure \ref{pan}. It can be infered that when random testing strategy is applied, Figure \ref{p.a}, comparing to the non-random testing strategy, Figure \ref{p.a}, the critical testing intensity of the coresponding panels are lower in non-random testing. Also, it appears that speeding the test reporting, i.e., increasing $\omega$, does not significantly lower $\Rnum$ when awaiting people follow the isolation perfectly, i.e., $\eta_w$ is closer to 0. However, speeding the test reporting reduces the epidemy more when the awaiting people less follow the isolation, i.e., $\eta_w$ is closer to 1.

@@

The next generation matrix is $G = F V^{-1}$, where matrix $F$ represents the inflow of new infection to the infected compartments and matrix $V$ represents the flow in the infected compartments when the population is totally susiptable. For model, \crefrange{eq1}{eq12}, $F$ and $V$ are

>>>>

\jd{There is a gap here between unravelling data and what we actually do.}
\bmb{As JD suggests, the focus of the paper is not on unraveling data (although that is something we \emph{could} do, and are trying to do, with this modeling framework as well). I would say that the ``bullet points'' to start the paper are (1) the processes driving the observed dynamics of the COVID-19 epidemic include both epidemiological processes (infection, recovery, etc.) and testing processes (``processes'' is repeated here, could rephrase); (2) in addition to determining what is actually observed (case reports), testing processes can also feed back to affect the epidemiological dynamics through isolation and quarantine; (3) your last sentence in the \P\ above.}

@@

\bmb{It's fine for this \P\ to be about testing, but the topic sentence still doesn't say what the paragraph is trying to do. You should probably give the general idea of how isolation and quarantine affect epi dynamics (i.e. individuals who in the process of testing may partially or fully self-isolate, and individuals who have tested positive are highly likely to self-isolate), and say that your goal is to figure out the effects of isolation and testing: when can isolation and testing control an epidemic? You might even want to do this in the \emph{previous} paragraph. Then you can say that since isolation applies to people who have been tested, that the epi effects will depend on \emph{who gets tested}, i.e. how tests are distributed across compartments, or for a given testing intensity how many people in each epidemiological compartment get tested. Then you can go through some examples, i.e. start by saying that in the case where tests are randomly assigned within the population (as might happen, for example, when disease surveillance is the purpose of testing), the tests will be weighted equally across all compartments \ldots}

It is often the purpose of testing, apart from the testing infrastructure, that dictates/defines \ali{[better word?]} the testing strategy. There are at least four different, and sometimes conflicting, purposes of testing: (i) diagnosis and treatment, (ii) screening people for access to things like flights, surgery, LTCs etc, (iii) surveillance, and (iv) mitigation and control of disease spread.

Further, the testing strategies can be grouped into two categories including random testing across the whole population and Non-random/biased testing. Random testing is the simplest form of testing strategy and is typically carried out for the purpose of surveillance. It is notable that, by random testing we mainly refer to PCR test given that there is almost negligible virological testing for surveillance purposes. Furthermore, all of the accessible testing data is being used for one of the other three purposes rather than surveillance. These are non-random testing strategies with a different bias. Specifically, when the purpose is "(i) diagnosis and treatment" or "(iv) mitigation and control of disease spread", the bias is towards infectious individuals more strongly for (i) than (iv). When people are tested for "(ii) screening", the bias is towards the group-specific characteristics [I am making this up for now!]. For example, people who are getting on flights may/will be more mobile than people awaiting for a surgery and are probably going to stay in long-term care homes and consequently more isolated to begin with.

-Examples of random versus biased testing strategies;
Each of these strategies has been used as a tool to study COVID-19; see (refs) for group (i) and (refs) for group (ii). The advantage of non-random testing comparing to random testing has been investigated specifically during the COVID-19 pandemic (refs).

@@

We developed a deterministic model, \crefrange{eq1}{eq12}, which groups individuals based on disease status and testing status. The former includes Susceptible, Infectious and Recovered, thus SIR, and the latter categorizes people as \emph untested, waiting-for-\emph positive, waiting-for-\emph negative, or \emph confirmed positive. Also, two 'accumulator' compartments, $N$ and $P$, were incorporated in the model in order to collect cumulative reported negative or positive tests. The model and details of calculation of $\Rnum$ are presented in the Supplementary Materials. The flowchart model is as follows

@@

The classical SIR model is based on the following implicit assumptions (see a ref for details); well-mixed population, homogeneity of the population, exponentially distributed duration of infection and large population size (may be refer the reader to \cite{keeling2011modeling}). In addition to these standard assumptions, our model, \crefrange{eq1}{eq12}, assumes: (i) there is a single force of infection, $\Lambda$, that is, where a susceptible individual moves to does not depend on who infected it, (ii) $\eta_c \leq \eta_w$, i.e., the individuals awaiting test results have a higher transmission probability than the reported individuals, (iii) a perfectly specific test, $p_S=0$, for simplicity in analysis. Note that the latter assumption combined with the assumption that the total population was initially distributed between untested and awaiting for negative test, that is $S_u+S_n=0$, reduce the model to 10 equations with equations \ref{eq3} and \ref{eq4} are eliminated. Specifically, when the population is at steady states and the number of infected people is negligible, the assumption of perfect specificity results in $S_p=0$ in \cref{eq3} and $d S_c/dt=0$ from \cref{eq4}. Also, since we assumed $S_u+S_n=0$, initially $S_c=0$ and stays in this state.

\newgeometry{margin=1cm} % modify this if you need even more space
\begin{landscape}
\begin{table}[htp]
\centering
{\tiny %
\begin{tabular}{|c|ccc|} \hline
  Symbol & Description & Unit & Value \\ \hline
  $N_0$     & Total population size & people & $10^6$ \\ \hline
  $\omega$  & Rate of onward flow from the awaiting to reported or untested compartments  & 1/day & - \\ \hline
  $\gamma$ & Recovery rate & 1/day & 1/3 \\ \hline 
  $\rho$   & Per capita testing intensity & 1/day & 0.01 \\ \hline 
  $\eta_w$  & Relative probability of transmission for isolated awaiting individuals & - & - \\ \hline
  $\eta_c$  & Relative probability of transmission for isolated confirmed individuals & - & -  \\ \hline
  $\Lambda$ & Force of infection & 1/time & - \\ \hline
  $p_S$ & Probability of false positive for susceptible& - & 0 \\ \hline
  $p_I$ & Probability of being infected and tested positive & - & 1 \\ \hline
  $p_R$ & Probability of being recovered and tested positive & - & 0.5 \\ \hline
  $W_S, W_I, W_R$ & Relative testing weight & - & 
  \begin{minipage}[t]{0.21\columnwidth}%
 Random testing: $W_S=W_I=W_R=1$, \\Non-random testing: $W_S=0.3, W_I=W_R=1$
\end{minipage} \\ \hline

@@

\todo{Something needs to be said here about simulation part.} The analytical calculation of the next generation matrix, $G$, and the derivation of $\Rnum$ was carried out in Maple (ref) by using simple linear Algebra package. 
  We used \Rlogo for the simulation part which included using the explicit expression of $\Rnum$ \ref{R0} as a function of the underlying parameters, specifying the parameters a realistic range, calculating the corresponding $\Rnum$ and plotting the contours of $\Rnum$ by using ggplot package in \Rlogo. The related result is presented in Figure \ref{pan}, where  panel \eqref{p.a} represents the random testing, and panel \eqref{p.b} is representing non-random testing. The parameter values or ranges are as follows (I like using a table here, with the param, their units and values). It is notable that in the simulation part, the explicit expression of $\Rnum$ was used and not the approximation \eqref{eq:R0appr}. This simulation reflects the behavior of $\Rnum$ with respect to the selected parameters for two different testing strategies: (i) random testing, represented by all testing weights to be the same, $W_S=W_I=W_R$, and (ii) non-random testing, when testing weight are not equal. For simulation purposes we chose $W_S=W_I=W_R=1$ for random testing, and $W_S=0.3$ and $W_I=W_R=1$ for non-random testing strategy. Note that the critical contour of $\Rnum=1$ is plotted in solid line in Figure \ref{pan}.

@@

Our numerical simulation in support of the analytical results in Proposition \ref{prop1} are presented in Figure \ref{pan}. It can be inferred that when random testing strategy is applied, Figure \ref{p.a}, comparing to the non-random testing strategy, Figure \ref{p.a}, the critical testing intensity of the corresponding panels are lower in non-random testing. Also, it appears that speeding the test reporting, i.e., increasing $\omega$, does not significantly lower $\Rnum$ when awaiting people follow the isolation perfectly, i.e., $\eta_w$ is closer to 0. However, speeding the test reporting reduces the epidemic more when the awaiting people less follow the isolation, i.e., $\eta_w$ is closer to 1.

@@

The next generation matrix is $G = F V^{-1}$, where matrix $F$ represents the inflow of new infection to the infected compartments and matrix $V$ represents the flow in the infected compartments when the population is totally susceptible. For model, \crefrange{eq1}{eq12}, $F$ and $V$ are
